Large-scale feature selection with Gaussian mixture models for the classification of high dimensional remote sensing images

### 2017-01-12 paper.tex
improve english

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available Earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched recently and 2 Terabytes of data will be released every day~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has also released thousands of images. Such high volume of Earth observation data provides accurate information of land state and functions, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

@@

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approaches (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to  users/developpers through  several
modules ``\emph{ready to use}''.

@@

There  are  two  main  strategies  to  reduce  dimension~\cite[Chapter
1]{Guyon:2006:FEF:1208773}: features extraction and feature selection.
Feature extraction means reformulate  and summarize the information by
creating new features in combining  the existing ones, it is sometimes
referred to as \emph{feature construction}.  Linear combination of the
initial features  can be extracted using  Principal Component Analysis
(PCA)~\cite{jimenez1998supervised}  or Independent  Component Analysis
\cite{villa2011hyperspectral}.  Supervised extraction  method has also
been investigated  such as Fischer discriminant  analysis and decision
boundary feature  extraction~\cite{book:landgrebe}.  To  the contrary,
features selection  extracts a subset of  existing features identified
as  the most  relevant  by a  given criterion.   This  subset has  the
additional advantage to  be much more understandable  for the end-user
than those constructed by a (non-)linear combination.

Features selection/extraction algorithms can be divided into three classes. The first class is called \emph{filter methods}. They select features independently of the classifier. Features are ranked according to some statistical measures, \emph{e.g.}, correlation or independence. For example,  PCA is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. Correlation between bands has been explored for feature selection in hyperspectral data~\cite{demir2008phase}. In general, these methods are fast and do not depend on any classifier. But they do not take into account the properties of the chosen classifier and do not optimize directly the classification accuracy.

The second class are known as \emph{wrapper methods}. They search for the best subset of variables for a given learning model. Since exhaustive searches are too expensive in terms of processing time, several sub-optimal search strategies have been designed, mainly iterative forward or backward search~\cite{whitney1971direct,marill1963effectiveness} or a combination of both~\cite{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model and to a particular learning problem. On the other hand, as they require the training of multiple models to test various set of variables, they are more time consuming.

@@

The third class corresponds to the \emph{embedded methods}. They do not separate the features selection process from the learning algorithm and allow interaction between the two processes. A popular embedded method is the \emph{Random Forest}. Embedded methods also exist for other models, \emph{e.g.} SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

Despite a large diversity of methods, feature selection algorithms usually do not scale well with the number of pixels to be processed~\cite{fauvel2015fast}. The training computational load is too important to compensate the reduced prediction computational load. Hence, feature selection is not widely used in operational situation. However, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

\rev{The contribution of the paper is a extension of the forward feature selection method proposed in~\cite{fauvel2015fast}. An smart implementation of the feature selection updates rules are presented in order to perform efficiently on large amount of data. They are based on linear algebra on block matrix applied to the covariance matrix of the conditional class density function.  Furthermore, a floating version of the algorithm is proposed and evaluated. Several \emph{goodness of fit} criteria are proposed to handle unbalanced training set, extending the conventional \emph{overall accuracy} measure.  Finally, the developed algorithm is made available to the scientific community as a remote module of the Orfeo Toolbox~\cite{christophe2008orfeo}.}

@@

The     remaining    of     the     article     is    organized     as
follows.   Section~\ref{sec:gmm-hd}  presents   GMM  classifiers   and
problems  related to  high-dimensional  feature  spaces. The  features
selection methods are detailed in Section~\ref{sec:selection}. Then an
efficient         implementation        is         presented        in
Section~\ref{sec:implementation}.  Experimental  results on  two  real
high dimensional data-set are given Section~\ref{sec:test}. Conclusion
and perspectives conclude the paper in Section~\ref{sec:conclusion}.

@@

    For mixture models, it is assumed  that a given sample $\mathbf{x}$ is
    the realization  of a  random vector which  distribution is  a mixture
    (convex     combination)     of      several     class     conditioned
    distribution~\cite{Fraley00model-basedclustering}:
    \begin{align}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{align}
    where $\pi_c$ is  the prior, \emph{i.e.}, the  proportion of class
    $c$  and  $f_c$  a  parametric density  function  control by
    $\theta$.

    Among the possible parametric model,  the Gaussian one is the most
    used~\cite{bouveyron2014model}.   It assumes  that each  $f_c$ is,
    conditionally  to  $c$,  a  Gaussian  distribution  of  parameters
    $\boldsymbol{\mu}_c$    and    $\boldsymbol{\Sigma}_c$:
    \begin{multline}
        f_c(\mathbf{x}|\boldsymbol{\mu}_c, \boldsymbol{\Sigma}_c) = \\ \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{multline}

@@

    There are two major solutions to this problem. First option is to stabilize the inversion of the covariance matrices. Some methods investigate the use of constraints on the direct problem. Reynolds \emph{et al.} \cite{reynolds1995robust} proposed to use diagonal covariance matrices. It is also possible to force the diagonal element to be higher than a given value by maximizing the GMM likelihood \cite{hathaway1985constrained}. Celeux and Govaert \cite{celeux1995gaussian} suggest to use equality constraints between coefficients of the covariance matrix in a parsimonious cluster-based GMM framework. Others papers propose to work on the inverse problem. A classical method is to use a regularization method as the well-known ridge regularization \cite{hoerl1970ridge}. A ridge regularization aims to stabilize the inversion by replacing the covariance matrix $\boldsymbol{\Sigma}_c$ by $\boldsymbol{\Sigma}_c + \tau \mathbf{I}$ where $\tau$ is a positive parameter and $\mathbf{I}$ the identity matrix. Jensen \emph{et al.} \cite{jensen2008regression} propose a different approach using a sparsity approximation to inverse the covariance matrix.

    Second option is to reduce the dimension. Features extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented.

@@

The feature selection method proposed in this work is a wrapper method associated to GMM models. Two elements are needed to set up a wrapper method:
\begin{enumerate}
\item A function that ranks the features according to some good classification or class separability criteria,
\item A search strategy to optimize the function.
\end{enumerate}

Section~\ref{sec:criterion} describes the different criterion used in this work and two search strategies are discussed in~\ref{sec:selection:method}.

    \subsection{Criterion function}
    \label{sec:criterion}
    The criterion evaluates how a given model built with a subset of features performs for the classification task. It can be an estimation of the correct classification or a measure of separability/similarity between class distribution. The former are in general more demanding in terms of processing time than the later.

@@

        \item \emph{The mean F1 score}  (F1mean) is the average of the
          F1 score  for each class  and the  F1 score is  the harmonic
          mean of  the precision  (number of  True Positive  over True
          Positive plus False Positive) and the recall (number of True
          Positive over True Positive plus False Negative)~\cite{powers2011evaluation}.
        \end{itemize}
        High  value  of  theses  indices  correspond  to  an  accurate
        classification.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions is  quantified using
        divergence  measure~\cite{opac-b1097517}. Contrary  to measure
        of correct  classification, divergences are  computed directly
        on  the  trained  model,  with  no  need  of  cross-validation
        estimation.  Two particular divergences are used in this work:
        the     \emph{Kullback-Leibler}     divergence     and     the
        \emph{Jeffries-Matusita}  distance.   The advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculus which  is a  major problem  when
        dealing with high-dimensional data.

@@

\section{Efficient implementation}
\label{sec:implementation}
The most  demanding part  of the  algorithm is  the evaluation  of the
criterion  for  all   the  remaining  variables  (see   lines  5-7  in
Algorithm~\ref{alg:sfs}). Calculus are based on linear algebra, and the
numerical  complexity is  on average  $O(d^3)$.  Furthermore,  for the
\emph{accuracy}-type  criterion the  complexity  is  augmented by  the
cross-validation procedure.

@@

    To deal with this limitation, the choice has been made to perform an eigendecomposition of the covariance matrix $\boldsymbol{\Sigma}^{(k)}_c$:

>>>>

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available Earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regular (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched recently and 2 Terabytes of data will be released every day~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has also released thousands of images. Such high volume of Earth observation data provides accurate information of land state and functions, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

@@

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approaches (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to  users/developpers through  several ``\emph{ready to use}'' modules.

@@

There  are  two  main  strategies  to  reduce  dimension~\cite[Chapter
1]{Guyon:2006:FEF:1208773}: feature extraction and feature selection.
Feature extraction means reformulate  and summarize the information by
creating new features in combining  the existing ones, it is sometimes
referred to as \emph{feature construction}.  Linear combination of the
initial features  can be extracted using  Principal Component Analysis
(PCA)~\cite{jimenez1998supervised}  or Independent  Component Analysis
\cite{villa2011hyperspectral}.  Supervised extraction  method has also
been investigated  such as Fischer discriminant  analysis and decision
boundary feature  extraction~\cite{book:landgrebe}.  To  the contrary,
feature selection  extracts a subset of  existing features identified
as  the most  relevant  by a  given criterion.   This  subset has  the
additional advantage to  be much more understandable  for the end-user
than those constructed by a (non-)linear combination.

Feature selection/extraction algorithms can be divided into three classes. The first class is called \emph{filter methods}. They select features independently of the classifier. Features are ranked according to some statistical measures, \emph{e.g.}, correlation or independence. For example,  PCA is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. Correlation between bands has been explored for feature selection in hyperspectral data~\cite{demir2008phase}. In general, these methods are fast and do not depend on any classifier. But they do not take into account the properties of the chosen classifier and do not optimize directly the classification accuracy.

The second class are known as \emph{wrapper methods}. They search for the best subset of variables for a given learning model. Since exhaustive searches are too expensive in terms of processing time, several sub-optimal search strategies have been designed, mainly iterative forward or backward search~\cite{whitney1971direct,marill1963effectiveness} or a combination of both~\cite{somol1999adaptive}. The advantage of such methods compared to filter methods is that they are dedicated to a particular model and to a particular learning problem. On the other hand, as they require the training of multiple models to test various set of variables, they are more time consuming.

@@

The third class corresponds to the \emph{embedded methods}. They do not separate the feature selection process from the learning algorithm and allow interactions between the two processes. A popular embedded method is the \emph{Random Forest}. Embedded methods also exist for other models, \emph{e.g.} SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

Despite a large diversity of methods, feature selection algorithms usually do not scale well with the number of pixels to be processed~\cite{fauvel2015fast}. The training computational load is too important to compensate the reduced prediction computational load. Hence, feature selection is not widely used in operational situations. However, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

\rev{The contribution of this paper is a extension of the forward feature selection method proposed in~\cite{fauvel2015fast}. A smart implementation of the feature selection update rules are presented in order to perform efficiently on large amount of data. The rules use on linear algebra on block matrices applied to the covariance matrix of the conditional class density function.  Furthermore, a floating version of the algorithm is proposed and evaluated. Several \emph{correctness of fit} criteria are proposed to handle unbalanced training sets, extending the conventional \emph{overall accuracy} measure.  Finally, the developed algorithm is made available to the scientific community as a remote module of the Orfeo Toolbox~\cite{christophe2008orfeo}.}

@@

The     remaining    of     the     article     is    organized     as
follows.   Section~\ref{sec:gmm-hd}  presents   GMM  classifiers   and
problems  related to  high-dimensional  feature  spaces. The  feature
selection methods are detailed in Section~\ref{sec:selection}. Then, an
efficient         implementation        is         presented        in
Section~\ref{sec:implementation}.  Experimental  results on  two  real
high dimensional data-sets are given Section~\ref{sec:test}. Conclusion
and perspectives conclude the paper in Section~\ref{sec:conclusion}.

@@

    For mixture models, it is assumed  that a given sample $\mathbf{x}$ is
    the realization  of a  random vector which  distribution is  a mixture
    (convex     combination)     of      several     class     conditioned
    distributions~\cite{Fraley00model-basedclustering}:
    \begin{align}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{align}
    where $\pi_c$ is  the prior, \emph{i.e.}, the  proportion of class
    $c$  and  $f_c$  a  parametric density  function  controlled by
    $\theta$.

    Among the possible parametric models,  the Gaussian one is the most
    used~\cite{bouveyron2014model}.   It assumes  that each  $f_c$ is,
    conditionally  to  $c$,  a  Gaussian  distribution  of  parameters
    $\boldsymbol{\mu}_c$    and    $\boldsymbol{\Sigma}_c$:
    \begin{multline}
        f_c(\mathbf{x}|\boldsymbol{\mu}_c, \boldsymbol{\Sigma}_c) = \\ \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{multline}

@@

    There are two major solutions to this problem. The first option is to stabilize the inversion of the covariance matrices. Some methods investigate the use of constraints on the direct problem. Reynolds \emph{et al.} \cite{reynolds1995robust} proposed to use diagonal covariance matrices. It is also possible to force the diagonal element to be higher than a given value by maximizing the GMM likelihood \cite{hathaway1985constrained}. Celeux and Govaert \cite{celeux1995gaussian} suggest to use equality constraints between coefficients of the covariance matrix in a parsimonious cluster-based GMM framework. Other papers propose to work on the inverse problem. A classical method is to use a regularization method as the well-known ridge regularization \cite{hoerl1970ridge}. A ridge regularization aims to stabilize the inversion by replacing the covariance matrix $\boldsymbol{\Sigma}_c$ by $\boldsymbol{\Sigma}_c + \tau \mathbf{I}$ where $\tau$ is a positive parameter and $\mathbf{I}$ the identity matrix. Jensen \emph{et al.} \cite{jensen2008regression} propose a different approach using a sparsity approximation to inverse the covariance matrix.

    The second option is to reduce the dimension. Feature extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented.

@@

The feature selection method proposed in this work is a wrapper method associated to GMM models. Two elements are needed to set up a wrapper method:
\begin{enumerate}
\item A function that ranks the features according to some good classification or class separability criterion,
\item A search strategy to optimize the function.
\end{enumerate}

Section~\ref{sec:criterion} describes the various criteria used in this work and two search strategies are discussed in~\ref{sec:selection:method}.

    \subsection{Criterion function}
    \label{sec:criterion}
    The criterion evaluates how a given model built with a subset of features performs for the classification task. It can be an estimation of the correct classification or a measure of separability/similarity between class distributions. The former are in general more demanding in terms of processing time than the later.

@@

        \item \emph{The mean F1 score}  (F1mean) is the average of the
          F1 score  for each class  and the  F1 score is  the harmonic
          mean of  the precision  (number of  True Positive  over True
          Positive plus False Positive) and the recall (number of True
          Positive over True Positive plus False Negative)~\cite{powers2011evaluation}.
        \end{itemize}
        High  values  of  theses  indices  correspond  to  an  accurate
        classification.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions can be  quantified using
        divergence  measures~\cite{opac-b1097517}. Contrary  to measures
        of correct  classification, divergences are  computed directly
        on  the  trained  model,  with  no  need  of  cross-validation
        estimation.  Two particular divergences are used in this work:
        the     \emph{Kullback-Leibler}     divergence     and     the
        \emph{Jeffries-Matusita}  distance.   The advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculations which  is a  major problem  when
        dealing with high-dimensional data.

@@

\section{Efficient implementation}
\label{sec:implementation}
The most  demanding part  of the  algorithm is  the evaluation  of the
criterion  for  all   the  remaining  variables  (see   lines  5-7  in
Algorithm~\ref{alg:sfs}). Calculations are based on linear algebra, and the
numerical  complexity is  on average  $O(d^3)$.  Furthermore,  for the
\emph{accuracy}-type  criterion the  complexity  is  augmented by  the
cross-validation procedure.

@@

    To deal with this limitation, the choice has been made to perform an eigenvalues decomposition of the covariance matrix $\boldsymbol{\Sigma}^{(k)}_c$:
